An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-04-25 EconGrowth.Rnw
Now including years of analysis in the countries table.

\usepackage{soul} %Provides strikethrough text
\usepackage{float} %Allows precise positioning of tables and figures within the text
\usepackage{amsmath} %Allows \begin{equation*} \end{equation*} for unnumbered equations

@@

The literature shows three quantifications of energy ($e$) used in production functions: thermal energy ($q$), exergy ($x$), and useful work ($u$). Thermal energy ($q$) is the heat content of an energy carrier. Exergy ($x$) is the maximum useful work that can be extracted from an energy carrier by a process that brings the system into equilibrium with a heat reservoir (typically the surrounding environment). Useful work ($u$) includes the direct energy that goes into an economy, including work from machines, human muscles, and animals. Most energy economists have used thermal energy ($q$) for the energy factor of production ($e$) in energy-augmented production functions, including \citet{Stern:2012ey}, \citet{Froling:2009vo}, \citet{Serrenho:2010vq}, and \citet{Kummel:1980wx,Kummel:1982vy,Kummel:1985vz,Kummel:1989tf,Kummel:2002tx,Kummel:2010vz}. \citet{Warr:2010ga} introduced the idea of using exergy ($x$) and useful work ($u$) in a production function to account for efficiency gains in energy production and consumption over the last century. We know of no statistically-rigorous comparisons of the predictive ability of each type of energy quantification in the various types of production functions discussed in Section \ref{sec:Prod_Func_Forms}.

@@

@@

In this paper, we evaluate

@@

In this paper we analyze economies at all stages of economic growth. We selected at least two economies from each of the following categories: the developed economies, which are economies that are technologically advanced with high standards of living; developing economies, which experience rapid economic growth but are still subject to low standards of living; the petrol economies, which depend on petroleum as a substantial source of GDP; and the Malthusian economies, which maintain living standards near subsistence levels. We analyzed the developed economies of the United States, the United Kingdom, and Japan for the period 1980-2011. We analyze two developing economies for 1991-2011: South Africa and China. The petrol economies analyzed for 1991-2011 are Iran and Saudi Arabia. Finally, the Malthusian economies analyzed from 1991-2011 are Tanzania and Zambia.

@@

<<Historical_Lattice_Graph, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.pos="H", fig.cap="Historical data: Indexed GDP ($y$), capital stock ($k$), labor ($l$), thermal energy ($q$), exergy ($x$), and useful work ($u$).", eval=TRUE>>=
createHistoricalLatticeGraph()
@

>>>>

\usepackage{soul} %Provides strikethrough text
\usepackage{float} %Allows precise positioning of tables and figures within the text
\usepackage{amsmath} %Allows \begin{equation*} \end{equation*} for unnumbered equations
\usepackage{multirow} % To create multirow tables.

@@

The literature shows three quantifications of energy ($e$) used in energy-augmented production functions: thermal energy ($q$), exergy ($x$), and useful work ($u$). Thermal energy ($q$) is the heat content of an energy carrier. Exergy ($x$) is the maximum useful work that can be extracted from an energy carrier by a process that brings the system into equilibrium with a heat reservoir (typically the surrounding environment). Useful work ($u$) includes the direct energy that goes into an economy, including work from machines, human muscles, and animals. Most energy economists have used thermal energy ($q$) for the energy factor of production ($e$) in energy-augmented production functions, including \citet{Stern:2012ey}, \citet{Froling:2009vo}, \citet{Serrenho:2010vq}, and \citet{Kummel:1980wx,Kummel:1982vy,Kummel:1985vz,Kummel:1989tf,Kummel:2002tx,Kummel:2010vz}. \citet{Warr:2010ga} introduced the idea of using exergy ($x$) and useful work ($u$) in a production function to account for efficiency gains in energy production and consumption over the last century. We know of no statistically-rigorous comparisons of the predictive ability of each type of energy quantification in the various types of production functions discussed in Section \ref{sec:Prod_Func_Forms}.

@@

In this paper, we evaluate example economies from four significant groups: Developed, BRICS, Petrol, and LDCs.\footnote{Lesser-Developed Countries}

In this paper we analyze at least two economies from each of the following categories: developed economies (technologically advanced with high standards of living), BRICS\footnote{Brazil, Russia, India, China, and South Africa} economies (rapid economic growth with standards of living below the developed economies), petrol economies (petroleum is a substantial source of GDP); and LDCs\footnote{Lesser Developed Countries} (living standards near subsistence levels). Table \ref{tab:Economy_Types} shows the example countries, their ISO 3166-1 alpha-2 codes, and the years of analysis for each economy.

\begin{table} \caption{Economy types and examples evaluated in this paper.} \label{tab:Economy_Types} 
  \begin{center}
    \begin{tabular}{r c c l l} 
      \multirow{3}{*}{Developed} & \includegraphics[height=0.11in]{flag/flag_US.png} & US & USA & \multirow{3}{*}{1980-2011}\\
                                 & \includegraphics[height=0.11in]{flag/flag_UK.png} & UK & United Kingdom & \\
                                 & \includegraphics[height=0.11in]{flag/flag_JP.png} & JP & Japan          & \\
      \hline
      \multirow{2}{*}{BRICS}     & \includegraphics[height=0.11in]{flag/flag_CN.png} & CN & China & \multirow{2}{*}{1990-2011}        \\
                                 & \includegraphics[height=0.11in]{flag/flag_ZA.png} & ZA & South Africa & \\
      \hline
      \multirow{2}{*}{Petrol}    & \includegraphics[height=0.11in]{flag/flag_SA.png} & SA & Saudi Arabia & \multirow{2}{*}{1990-2011} \\
                                 & \includegraphics[height=0.11in]{flag/flag_IR.png} & IR & Iran         &  \\
      \hline
      \multirow{2}{*}{LDC}       & \includegraphics[height=0.11in]{flag/flag_TZ.png} & TZ & Tanzania & \multirow{2}{*}{1990-2011}     \\
                                 & \includegraphics[height=0.11in]{flag/flag_ZM.png} & ZM & Zambia      & \\
    \end{tabular}
  \end{center}
\end{table}

@@

@@

@@

<<Historical_Lattice_Graph, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.pos="H", fig.cap="Historical data: Indexed GDP ($y$), capital stock ($k$), labor ($l$), thermal energy ($q$), exergy ($x$), and useful work ($u$).">>=
createHistoricalLatticeGraph()
@
